---
title: "Credit Risk via Lending Club"
author: " Modern Data Mining Group 39"
date: "2022/3/20"
geometry: margin=1in
output:
  html_document:
    code_folding: hide
    highlight: haddock
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
urlcolor: blue
always_allow_html: true
indent: true 
fontsize: 12pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 7, fig.height = 4)
if(!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, glmnet, car, data.table,summarytools,rpart.plot,r.part)   #add your packages here
knitr::opts_knit$set(root.dir = '/Users/ksun/Downloads/stat571/stat571-team-39/hw4')
```



# EDA

```{r}
lending_data <- fread("LendingClubCase/LoanStats_07_11_Clean.csv")
lending.data.pre = lending_data[,-c("issue_d","funded_amnt","funded_amnt_inv","total_pymnt","total_pymnt_inv","total_rec_prncp","total_rec_int","total_rec_late_fee","recoveries","collection_recovery_fee","last_pymnt_d","last_pymnt_amnt","last_credit_pull_d")]
possible.factor = c("loan_amnt","int_rate","grade","home_ownership","annual_inc","addr_state","dti","verification_status","delinq_2yrs","total_acc","pub_rec","revol_bal","pub_rec_bankruptcies")
```

## default vs. grades

As shown in the barplot below, the percentage of fully-paid loans decreases steadily from Grade A to Grade G. Approximately 6% of Grade A loans defaulted whereas 31% of Grade G loans defaulted. 

```{r}
lending.sum = lending_data[order(grade),.(
  fully.paid.prop = length(which(loan_status == "Fully Paid"))/length(loan_status)
),by=grade]

ggplot(data = lending.sum,aes(x=grade,weight=fully.paid.prop))+
  geom_bar(fill = "Darkblue")+
  ylab("prop of fully paid loans")+
  theme_bw()
```

## default vs. annual income

For simplicity we start with a simple question: How does loan status relate to the annual income of the individual?

```{r}
income = lending_data %>% group_by(loan_status) %>% summarise(mean(annual_inc))
```

We observe that defaulted loans were held by people with an average income of 62,638, while non-defaulted loans were held by people with an average income of 70,082. Therefore we can see that defaults are negatively correlated with low income.

## default vs. interest rate

```{r}
ggplot(data = lending_data,aes(x=int_rate,y=loan_status))+
  geom_point(position = "jitter")+
  theme_bw()
```





# Multiple Logistic Regression and Classification

# Decision Tree

```{r}
tree_data <- lending_data %>% select(c(loan_status, dti, delinq_2yrs,  inq_last_6mths, revol_util, revol_bal, open_acc, pub_rec, total_acc, pub_rec_bankruptcies))
# tree_data$loan_status <- as.factor(tree_data$loan_status)
tree_data$loan_status = tree_data$loan_status != "Fully Paid"
```

Create a function that splits data into train and test sets

```{r}
create_train_test <- function(data, size = 0.8, train = TRUE) {
    n_row = nrow(data)
    total_row = size * n_row
    train_sample <- 1: total_row
    if (train == TRUE) {
        return (data[train_sample, ])
    } else {
        return (data[-train_sample, ])
    }
}
```

Split the Data into train/test set
```{r}
data_train <- create_train_test(tree_data, 0.8, train = TRUE)
data_test <- create_train_test(tree_data, 0.8, train = FALSE)
dim(data_train)
```
Verify that the randomization process is correct
```{r}
prop.table(table(data_train$loan_status))
prop.table(table(data_test$loan_status))
```
```{r}
MCE = function(prediction,actual){
  False_negative = sum(prediction[actual == TRUE] != TRUE)
  False_positive = sum(prediction[actual != TRUE] == TRUE)
  return((False_negative*2.2 + False_positive)/length(actual))
}
```


```{r}
#Baseline Model predicting all as would not default
MCE(rep(FALSE,nrow(tree_data)), tree_data$loan_status)
```
```{r}
mod1 = glm(data = data_train,loan_status~.,family = "binomial")
```
```{r}
summary(mod1)
```

```{r}
#Confusion Matrix
table(predict(mod1,data_test,type = "response") > (1/2.2)/(1+(1/2.2)),
    data_test$loan_status)
```

```{r}
#Calculate MCE
MCE(predict(mod1,data_test,type = "response") > (1/2.2)/(1+(1/2.2)),
    data_test$loan_status)
```

```{r}
mod0_ROC = pROC::roc(as.integer(data_test$loan_status), predict(mod1,data_test,type = "response"))
pROC::auc(mod0_ROC)
```

```{r}
model_tree <- rpart(loan_status~., data = data_train, method = "class", cp=0.00096)
rpart.plot(model_tree)
```


```{r}
tree_ROC = pROC::roc(as.integer(data_test$loan_status), predict(model_tree,newdata = data_test,type = "prob")[,2] )
print(pROC::auc(tree_ROC))
```
```{r}
library(randomForest)
```
```{r}
model_rf = randomForest::randomForest(loan_status~., data_train, method = "class", mtry = 5)
```
```{r}
model_rf
```
```{r}
table(predict(model_rf,data_test,type = "response") > (1/2.2)/(1+(1/2.2)),
    data_test$loan_status)
```
```{r}
rf_ROC = pROC::roc(as.integer(data_test$loan_status), predict(model_rf,newdata = data_test,type = "response"))
print(pROC::auc(rf_ROC))
```

ii) Analyses

* Various appropriate statistical methods: e.g. glmnet (and/or trees, ignore this at the moment)
* Comparisons various models
* Final model(s)

iii) Conclusion

* Summarize results and the final model
* Caveats
* Final recommendations

Maintain a good descriptive flow in the text of your report. Use Appendices to display lengthy output. 

iii) Appendix
	
* All your R code (code without comments is no good!) if you are not using `rmd` format.
* Any thing necessary to keep but for which you don’t want them to be in the main report.

