---
title: "HW3"
author: "Liming Ning"
date: "2022/2/21"
geometry: margin=1in
output:
  html_document:
    code_folding: hide
    highlight: haddock
    number_sections: yes
    toc: yes
    toc_depth: 4
    toc_float: yes
urlcolor: blue
always_allow_html: true
indent: true 
fontsize: 12pt
---

```{r setup, include=F, cache=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F)
knitr::opts_knit$set(root.dir = 'G:/github/modern-data-mining/hw3/liming')
output_format <- ifelse(is.null(knitr::opts_knit$get("rmarkdown.pandoc.to")),
                        "text", knitr::opts_knit$get("rmarkdown.pandoc.to"))
options(scipen = 0, digits = 3)  # controls base R output
library(data.table)
library(reshape2)
library(readxl)
library(lubridate)
library(tidyverse)
library(cowplot)
library(kableExtra)
library(ggrepel)
library(skimr)
library(plotly)
library(RColorBrewer)
library(ggbiplot)
library(factoextra)
library(DescTools)
library(irlba)
library(ISLR)
library(lmtest)
library(stargazer)
library(GGally)
library(summarytools)
library(usmap)
Sys.setlocale(,"English")
```

# covid case

```{r}
# county-level socialeconomic information
county_data <- fread("data/covid_county.csv")
# county-level COVID case and death
covid_rate <- fread("data/covid_rates.csv")
# county-level lockdown dates
covid_intervention <- fread("data/covid_intervention.csv")

```

## understand the data

```{r,eval=F}
view(dfSummary(covid_rate),method = "render")
view(dfSummary(county_data),method = "render")
```

## covid case trend 

### three states, by day and state
```{r}
unique(covid_rate$State)
covid.3state.day = covid_rate[State %in% c("New York","Washington","Florida"),.(
  cum_cases = sum(cum_cases),
  cum_deaths = sum(cum_deaths),
  week = mean(week)
),by = .(date,State)]
covid.3state.day = covid.3state.day[order(list(State,date))]

covid.3state.day[,new_cases := c(NA,diff(cum_cases)),by = State]
covid.3state.day[,date := ymd(date)]
ggplot(covid.3state.day,aes(x=date,y=new_cases,group=State,color=State))+
  geom_line()+
  scale_x_date(date_breaks = "3 month",date_labels = "%Y-%m")+
  labs(title = "New Cases Trends for NY, WA and FL")+
  theme_bw()
```

The biggest problem: daily variability is extremely high, making the trends zigzagging. Maybe there is an innate difference among days in a week because people have different work and life style in different days. We may wish to smooth out these excessive noise by aggregating in the weekly level.

### (ii)

```{r}
# check
covid_rate[State == "New York",length(County),by = .(State,date)] # problem: some counties do not report cases in some days. New added counties cause problems

covid.weekend = covid_rate[,.(
  cum_cases_weekend = cum_cases[length(cum_cases)],
  cum_deaths_weekend = cum_deaths[length(cum_deaths)],
  TotalPopEst2019.weekend = TotalPopEst2019[length(TotalPopEst2019)]
), by = .(week,State,County)]
covid.weekend = covid.weekend[order(list(State,County,week))]
covid.weekend[,":="(new_cases = c(NA,diff(cum_cases_weekend)),
                                 new_deaths = c(NA,diff(cum_deaths_weekend))),
                           by = .(State,County)]
covid.week = covid.weekend[!is.na(new_cases)&!is.na(new_deaths),.(
  new_cases = sum(new_cases),
  new_deaths = sum(new_deaths),
  TotalPopEst2019 = sum(TotalPopEst2019.weekend)
),by = .(State,week)]
covid.week[,weekly_case_per100k := new_cases/TotalPopEst2019*100000]

# data.ma = covid_rate[State == "Massachusetts"] # pay attention: week 33 of MA

ggplot(covid.week,aes(x=week,y=weekly_case_per100k,group=State,color=State))+
  geom_line()+
  labs(title = "Spaghetti Plots, Weekly New Cases")+
  theme_bw()
```

### (iii)



### (iv)

```{r}
summary(covid_intervention)
# some examples

covid_intervention[STATE == "New York"]
covid_intervention_state = covid_intervention[substr(FIPS,nchar(FIPS)-2,nchar(FIPS)) == "000"]

# NY
ggplot(covid.3state.day[State == "New York"],aes(x=date,y=new_cases))+
  geom_line()+
  scale_x_date(date_breaks = "3 month",date_labels = "%Y-%m")+
  geom_vline(xintercept = covid_intervention_state[STATE == "NY","stay at home"][[1,1]],color = "red",lty = 5)+
  theme(title = "New Cases Trends for NY")+
  theme_bw()

# FL
ggplot(covid.3state.day[State == "Florida"],aes(x=date,y=new_cases))+
  geom_line()+
  scale_x_date(date_breaks = "3 month",date_labels = "%Y-%m")+
  geom_vline(xintercept = covid_intervention_state[STATE == "FL","stay at home"][[1,1]],color = "red",lty = 5)+
  labs(title = "New Cases Trends for FL")+
  theme_bw()

```

## covid death trend

### Monthly deaths per 100k heatmap
```{r}
# pay attention to numbers smaller than zero
covid_rate = covid_rate[order(FIPS,date)]
covid_rate[,":="(new_cases = c(NA,diff(cum_cases)),
                 new_deaths = c(NA,diff(cum_deaths)),
                 year = year(date),
                 month = month(date),
                year.month = round_date(date,"month")),by = FIPS]
covid.month = covid_rate[!is.na(new_cases)&!is.na(new_deaths),.(
  new_cases = sum(new_cases),
  new_deaths = sum(new_deaths),
  TotalPopEst2019 = TotalPopEst2019[length(TotalPopEst2019)]
),by = .(State,County,year,month)]
covid.month = covid.month[,.(
  new_cases = sum(new_cases),
  new_deaths = sum(new_deaths),
  TotalPopEst2019 = sum(TotalPopEst2019)
),by = .(State,year,month)]

covid.month[,monthly_death_per100k := new_deaths/TotalPopEst2019*100000]
```

```{r}
covid.death.plot.list = list()
setnames(covid.month,"State","state")
max_col = quantile(covid.month$monthly_death_per100k,1,na.rm = T)
min_col = quantile(covid.month$monthly_death_per100k,0,na.rm = T)
for (i in 2:12) {
  covid.death.plot.list[[i-1]] =
    plot_usmap(regions = "state",data = covid.month[year == 2020 & month == i],
               values = "monthly_death_per100k", exclude = c("Hawaii", "Alaska"),color = "black") + 
    scale_fill_distiller(
      palette = "Reds",direction = 1,
      name = "Number of New Covid Deaths per 100,000 People", 
      limits = c(min_col, max_col)) + 
    labs(title = paste0("New Covid Deaths, 2020-",i), subtitle = "Continental US States") +
    theme(legend.position = "right")
}
# ggplotly(covid.death.plot.list[[2]])

# plot_usmap(regions = "state",data = covid.month[year == 2020 & month == i],
#                values = "monthly_death_per100k", exclude = c("Hawaii", "Alaska"),color = "black") + 
#     scale_fill_gradient(
#       low = "white", high = "red", 
#       name = "Number of New Covid Deaths per 100,000 People", 
#       label = scales::comma) + 
#     labs(title = paste0("New Covid Deaths, 2020-",i), subtitle = "Continental US States") +
#     theme(legend.position = "right")


```

### (ii)

```{r}
# ggplotly

test.long = covid.month[year == 2020,.(month,state,monthly_death_per100k)]
# test = test.long %>%
#   pivot_wider(names_from = month,values_from = monthly_death_per100k) %>%
#   mutate(state = tolower(state))
# 
# map.wide = test %>% 
#   left_join(map_data("state"),by = c("state" = "region"))
# map.wide = data.table(map.wide)
# map = map.wide[order(order)] %>% 
#   pivot_longer(cols = 2:13,names_to = "month",values_to = "monthly_death_per100k") %>% 
#   mutate(month = as.numeric(month))
# map = data.table(map)
# map = map[order(list(month,order))]
# 
# plot.test = ggplot(map,aes(x=long,y=lat,group=group))+
#   geom_polygon(aes(fill = monthly_death_per100k))+
#   geom_path()+
#   scale_fill_distiller(palette = "Reds", direction = 1,
#                        name = "Num", 
#       limits = c(min_col, max_col))+
#   labs(title = paste0("New Monthly Covid Deaths per 100k"), subtitle = "Continental US States")+
#   theme_bw()
# ggplotly(plot.test) # error?

# plotly
abbr = unique(us_map(regions =
"states") %>% select(abbr, full)) 

plotly.data = covid.month[year == 2020,.(month,state,monthly_death_per100k)] %>%
  mutate(hover =  paste(state, '<br>', 
                        'new covid deaths', round(monthly_death_per100k, 3))) %>% 
  left_join(abbr,by = c("state" = "full"))

# give state boundaries a white border
l <- list(color = toRGB("white"), width = 2)
# specify some map projection/options
g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white'),
  exclude = c("")
)
plot_ge
fig <- plot_geo(plotly.data, locationmode = 'USA-states')
fig <- fig %>% add_trace(
    z = ~monthly_death_per100k, text = ~hover, locations = ~abbr,
    color = ~monthly_death_per100k, colors = 'Reds',
    zmin = min_col, zmax = max_col,frame = ~month
  )

fig <- fig %>% colorbar(title = "Monthly new covid deaths of state")
fig <- fig %>% layout(
    title = 'Monthly new covid deaths of state',
    geo = g,
    hoverlabel = list(bgcolor="white")
  )
fig
```

```{r,eval=F}
save.image("hw3_covid.RData")
```

```{r}
load("codes/hw3_covid.RData")
```


## covid factor

```{r}
county.covid = covid_rate[date == as.Date("2021-02-01"),.(
  FIPS,County,State,
  total_death_per100k = cum_deaths/TotalPopEst2019*100000
)] %>% 
  right_join(county_data, by = "FIPS") %>%
  mutate(log_total_death_per100k = log(total_death_per100k + 1))

county.covid.sub <- county.covid %>%
  select(log_total_death_per100k,State.x,County.x, FIPS, Deep_Pov_All, PovertyAllAgesPct, PerCapitaInc, UnempRate2019, PctEmpFIRE, PctEmpConstruction, PctEmpTrans, PctEmpMining, PctEmpTrade, PctEmpInformation, PctEmpAgriculture, PctEmpManufacturing, PctEmpServices, PopDensity2010, OwnHomePct, Age65AndOlderPct2010, TotalPop25Plus, Under18Pct2010, Ed2HSDiplomaOnlyPct, Ed3SomeCollegePct, Ed4AssocDegreePct, Ed5CollegePlusPct, ForeignBornPct, Net_International_Migration_Rate_2010_2019, NetMigrationRate1019, NaturalChangeRate1019, TotalPopEst2019, WhiteNonHispanicPct2010, NativeAmericanNonHispanicPct2010, BlackNonHispanicPct2010, AsianNonHispanicPct2010, HispanicPct2010, Type_2015_Update, RuralUrbanContinuumCode2013, UrbanInfluenceCode2013, Perpov_1980_0711, HiCreativeClass2000, HiAmenity, Retirement_Destination_2015_Update)

setnames(county.covid.sub,c("State.x","County.x"),c("state","county"))
```

```{r}
county.covid.sub[is.na(state)]
apply(is.na(county.covid.sub), 2, sum) # missing in state, county: all aggregate states, puerto rico, hawaii, alaska; bedford in Virginia has two obs, duplicated

county.covid.sub = na.omit(county.covid.sub) # -174
```

### lasso

```{r}
colnames(model.var)
model.var = model.matrix(log_total_death_per100k~.,data = county.covid.sub[,-c("FIPS","county")])[,-1]
# head(model.test)
death = county.covid.sub[,log_total_death_per100k]
fit1.cv = cv.glmnet(model.var,death,alpha = 1, nfolds = 10, intercept = T,
                    penalty.factor = c(rep(0,48),rep(1,ncol(model.var)-48))) # alpha: the para in elastic net
coef.min = coef(fit1.cv,s="lambda.1se") # more parsimonious usually. If use min, maybe there is noise
# coef(fit1.cv,s="lambda.1min")
nonzero.coef = coef.min[which(coef.min!=0),]
# plot(fit1.cv)
nonzero.var = names(nonzero.coef)[-1]
```

### fine tune the model

```{r}
data.selected = data.table(death,model.var[,which(colnames(model.var) %in% nonzero.var)])

fit.1se.lm = lm(death~.,data = data.selected)
summary(fit.1se.lm)
# colnames(data.selected)
fit.final.1 =  regsubsets(death~.,data = data.selected,method = "exhaustive",
                          nvmax = ncol(data.selected)-1,force.in = c(2:48),really.big = T) # compared to Arizona, anomaly in force.in
summary.fit.final.1 = summary(fit.final.1)
plot(summary.fit.final.1$bic)
opt.index = 11 # with bic

bic.var.select = summary.fit.final.1$which[opt.index,-1]
bic.var = names(bic.var.select)[which(bic.var.select)] 
# bic.var
final.expr = as.formula(paste("death", "~", paste(bic.var, collapse = "+"))) 
fit.final.2 = lm(final.expr,data = data.selected)
summary(fit.final.2) # all significant
```

Elderly: $Age65AnOlderPct2010$ has a significantly positive coefficient. 

Hispanic/Black: not in the regression after controlling for existing variables.

### diagnosis

```{r}
scatter.list = list()

for (i in 48:58) { # cannot do this. the plot is built only after it's invoked if in the loop? Use aes_string
  name = bic.var[i]
  scatter.list[[i-47]] = ggplot(data.selected,aes_string(x=name,y="death"))+
    geom_point()+
    geom_smooth(method = "lm")+
    xlab(name)+
    ylab("log.new.death")+
    theme_bw()
}

plot_grid(plotlist = scatter.list)


# residual plot
plot(fit.final.2,1)

# qq plot
plot(fit.final.2,2)
```

Seems not a good fit.